---
title: "Introduction to segtest Version 2.0"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro_2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1)
```

```{r setup}
library(segtest)
```

Segtest provides a suite of tests for segregation distortion for F1 populations of polyploids under various models of meiosis. You can use these tests for autopolyploids that exhibit full polysomic inheritance, allopolyploids that exhibit full disomic inheritance, or segmental allopolyploids that exhibit partial preferential pairing. Double reduction is (optionally) fully accounted for in tetraploids, and (optionally) partially accounted for (only at simplex loci) for higher ploidies. Some maximum proportion of outliers can be specified, and so these methods can accommodate moderate levels of double reduction at non-simplex loci. Offspring genotypes can either be known, or genotype uncertainty can be represented through genotype likelihoods. Parent data may or may not be provided, at your option. Parents can have different (even) ploidies, at your option.

Additional functions include those that generate gamete and genotype frequencies under different models of meiosis, functions that simulate genotype (log) likelihoods, and competing tests for segregation distortion.

The main functions are:

- `seg_multi()`: Run the likelihood ratio test for segregation distortion in parallel at many loci.
- `multidog_to_g`: Format the genotyping output from `updog::multidog()` to be compatible withe input of `seg_multi()`.
- `seg_lrt()`: Test for segregation distortion for any even ploidy.
- `gamfreq()`: Gamete frequencies.
- `gf_freq()`: Genotype frequencies of an F1 population of polyploids.
- `drbounds()`: Upper bounds on the double reduction rate(s) based on two different
   extreme models of meiosis.
- `simgl()`: Simulate genotype log-likelihoods given a vector of genotype counts.

# Gamete genotype frequencies

`gamfreq()` will generate gamete frequencies under different models of meiosis. `gf_freq()` will generate genotype frequencies under the same models by convolving the output from `gamfreq()`. We focus here on `gamfreq()` since `gf_freq()` has the same models, but specified for each parent.

For autopolyploids, specify `type = "polysomic"` and, optionally, the amount of double reduction via `alpha`. `alpha` is a vector of length `floor(ploidy / 4)` where element `i` is the probability a gamete has genotype `i`. The upper bounds for `alpha` can be found via `drbounds()`. E.g., for a parental octoploid with genotype 4 with no and moderate levels of double reduction:
```{r}
drbounds(ploidy = 8) ## DR bounds
gamfreq(g = 4, ploidy = 8, type = "polysomic") ## no DR
gamfreq(g = 4, ploidy = 8, alpha = c(0.1, 0.01), type = "polysomic") ## Some DR
```

For allopolyploids, the possible gamete frequencies can be found in `seg`. E.g., for a parental octoploid with genotype 4, the possible gamete frequencies are
```{r}
seg[seg$ploidy == 8 & seg$g == 4 & seg$mode %in% c("disomic", "both"), "p"]
```
Note that you also need to filter for the `mode` to be either `"disomic"` or `"both"` (both disomic and polysomic). The total number of possible allopolyploid distributions is `n_pp_mix()`.
```{r}
n_pp_mix(g = 4, ploidy = 8)
```
You can specify one of these distributions via a 1-of-3 vector. E.g.
```{r}
gamfreq(g = 4, ploidy = 8, gamma = c(1, 0, 0), type = "mix")
gamfreq(g = 4, ploidy = 8, gamma = c(0, 1, 0), type = "mix")
gamfreq(g = 4, ploidy = 8, gamma = c(0, 0, 1), type = "mix")
```

Segmental allopolyploids are mixtures of the possible allopolyploid segregation distributions. E.g., an equal mixture of the three for an octoploid with genotype 4 is
```{r}
gamfreq(g = 4, ploidy = 8, gamma = c(1, 1, 1)/3, type = "mix") 
```

At simplex, loci, there is only one possible allopolyploid segregation distribution:
```{r}
n_pp_mix(g = 1, ploidy = 8)
gamfreq(g = 1, ploidy = 8, gamma = 1, type = "mix")
n_pp_mix(g = 7, ploidy = 8)
gamfreq(g = 7, ploidy = 8, gamma = 1, type = "mix")
```
You can account for double reduction at these loci by including `beta`. The upper bound of which can be found via `beta_bounds()`.
```{r}
beta_bounds(ploidy = 8)
gamfreq(g = 1, ploidy = 8, gamma = 1, beta = 0.03, type = "mix")
gamfreq(g = 7, ploidy = 8, gamma = 1, beta = 0.03, type = "mix")
```

# Simulating data

Let's suppose we have some genotype frequencies we want to simulate individual data from:
```{r}
gf <- gf_freq(
  p1_g = 2, 
  p1_ploidy = 6,
  p1_gamma = c(0.7, 0.3), 
  p1_type = "mix",
  p2_g = 4,
  p2_ploidy = 6, 
  p2_gamma = c(0.5, 0.5), 
  p2_type = "mix")
plot(gf, type = "h")
```

To simulate genotype counts, just use `r`multinom()` from the `stats` package. Let's simulate data from 10 individuals.
```{r}
x <- c(rmultinom(n = 1, size = 10, prob = gf))
x
```
To simulate genotype (log) likelihoods, insert these genotype counts into `simgl()`.
```{r}
gl <- simgl(nvec = x)
gl
```

# Testing for segregation distortion

You can test for segregation distortion using `seg_lrt()`. E.g., let's test for it using the data (both known genotypes and genotype likelihoods) we simulated from the previous section:
```{r}
## With known genotypes
sout1 <- seg_lrt(x = x, p1_ploidy = 6, p2_ploidy = 6, p1 = 2, p2 = 4)
sout1$p_value
## With genotype likelihoods
sout2 <- seg_lrt(x = gl, p1_ploidy = 6, p2_ploidy = 6, p1 = 2, p2 = 4)
sout2$p_value
```

My recommendation is to always use the genotype log-likelihoods. But `seg_lrt()` allows for known genotypes, if that situation works best for you.

The default (`model = "seg"`) is to assume your organism is a segmental allopolyploids, and to account for possible double reduction at simplex loci. But you should absolutely use other models if you have more information on your organism:

- `"seg"`: General segmental allopolyploid with possible double reduction at simplex loci.
- `"allo_pp"`: Segmental allopolyploid with complete bivalent pairing (no double reduction).
- `"allo"`: Pure allopolyploid (disomic inheritance).
- `"auto"`: Pure autopolyploid with complete bivalent pairing (no double reduction).
- `"auto_dr"`: Pure autopolyploid with possible multivalent pairing (some double reduction).
- `"auto_allo"`: Same null hypothesis as in polymapR.

We allow for some non-valid genotypes via the `ob` argument. This is the upper bound on the proportion of outliers. By default, this is set to 0.03. You can set this to `0` (or set `outlier = FALSE`) if you want any outliers to indicate segregation distortion.

Make sure that the log-likelihoods are base e. If they are base 10, you'll get the wrong p-value:
```{r}
gl10 <- gl / log(10)
seg_lrt(x = gl10, p1_ploidy = 6, p2_ploidy = 6, p1 = 2, p2 = 4)$p_value
```

Don't mess with the technical arguments (`ntry`, `opt`, `optg`, `df_tol`). These have to do with the optimization and how to approximate the degrees of freedom of the test. Except possibly `ntry`. You could increase that if you are seeing weird results. But then let me know, because I haven't seen any bad behavior with `ntry = 3`.
