---
title: "Parallel Testing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parallel Testing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1)
```

```{r setup}
library(segtest)
```

# Introduction

We provide a walk-through of how to run tests for segregation distortion at many loci in parallel.

# Analysis

Data sets `ufit`, `ufit2`, and `ufit3` contain the genotyping output of `updog::multidog()` using three different models.

- `ufit`: Uses the `norm` model without specifying who the parents are.
- `ufit2`: Uses the `f1pp` model, specifying the parents.
- `ufit3`: Uses the `f1` model, specifying the parents.

You can convert this genotyping output to what `seg_multi()` expects using `multidog_to_g()`.

If you did *not* use the `f1pp` or `f1` models, use ether the `all_gl` (to run tests using genotype log-likelihoods) or `all_g` (to run tests assuming genotypes are known) options, and you must specify the ID's of the parents.

```{r}
o1 <- multidog_to_g(ufit, type = "all_g", p1 = "indigocrisp", p2 = "sweetcrisp")
o2 <- multidog_to_g(ufit, type = "all_gl", p1 = "indigocrisp", p2 = "sweetcrisp")
```

If you *did* use the `f1pp` or `f1` models, use either the `off_gl` (to run tests using genotype log-likelihoods) or `off_g` (to run tests assuming genotypes are known) options.

```{r}
o3 <- multidog_to_g(ufit2, type = "off_g")
o4 <- multidog_to_g(ufit2, type = "off_gl")
o5 <- multidog_to_g(ufit3, type = "off_g")
o6 <- multidog_to_g(ufit3, type = "off_gl")
```

We would recommend *always* using genotype log-likelihoods. But the option is there for known genotypes.

Parallelization support is provided through the future package. You specify how to implement parallelization using `future::plan()`. Then you run `seg_multi()`. Then you shut down the parallelization with `future::plan()`. The most common plan is `future::multisession()`, where you specify the number of parallel processes with the `workers` argument. You can get the maximum number of workers via `future::availableCores()`

```{r}
future::availableCores()
```

So a typically workload looks something like this:
```{r}
future::plan(future::multisession(workers = 2))
mout1 <- seg_multi(
  g = o2$g, 
  p1 = o2$p1,
  p2 = o2$p2, 
  p1_ploidy = 4, 
  p2_ploidy = 4,
  ret_out = TRUE)
future::plan(future::sequential())
```

The output is a data frame. The most important parts of this are the `snp` and `p_value` columns.
```{r}
mout1[c("snp", "p_value")]
```


It looks like only SNP *12_31029646* is in possible segregation distortion. Let's compare the expected frequencies (from the `q0` column) against the observed modes.
```{r}
round(mout1$q0[mout1$snp == "12_31029646"][[1]], digits = 4)
o1$g["12_31029646", ] / sum(o1$g["12_31029646", ])
```
So SNP *12_31029646* likely got flagged because there are too many individuals with a genotype of 2, and not enough with a genotype of 0.

If we would have run `outlier = FALSE`, then SNP *2_20070837* would also have been flagged for possible segregation distortion
```{r}
future::plan(future::multisession(workers = 2))
mout2 <- seg_multi(
  g = o2$g,
  p1 = o2$p1, 
  p2 = o2$p2, 
  p1_ploidy = 4, 
  p2_ploidy = 4,
  outlier = FALSE)
future::plan(future::sequential())
```
```{r}
mout2[, c("snp", "p_value")]
```

Let's look at the posterior mode genotypes of SNP 2_20070837:
```{r}
o1$g["2_20070837", ]
o1$p1["2_20070837"]
o1$p2["2_20070837"]
```
So SNP 2_20070837 likely got flagged because of two individuals that are very likely a genotype of 3, which is impossible if the parents have genotypes 0 and 2. Including `outlier = TRUE` previously made it so that we would ignore this discrepancy (up to a point). You can get the posterior probability that an individual is an outlier if you set `ret_out = TRUE` (which I previously did).

```{r}
graphics::plot(
  mout1$outprob[mout1$snp == "2_20070837"][[1]], 
  ylim = c(0, 1),
  ylab = "Outlier Probability")
```

# `multi_lrt()`

The older function `multi_lrt()`, only provides support for tetraploids. You can still use it instead of `seg_multi()` for tetraploids. 
```{r}
future::plan(future::multisession(workers = 2))
mout3 <- multi_lrt(
  g = o2$g,
  p1 = o2$p1, 
  p2 = o2$p2
)
future::plan(future::sequential())
```
It's a little faster, but it cannot account for outliers, so it will mostly be the same as setting `outlier = FALSE`. 
```{r}
plot(
  mout2$p_value,
  mout3$p_value,
  xlab = "seg_multi(outlier = FALSE)", 
  ylab = "multi_lrt()")
abline(0, 1, lty = 2, col = 2)
```
The one discrepancy above is caused by the new way we calculate the degrees of freedom:
```{r}
i <- which.max(abs(mout2$p_value - mout3$p_value))
mout2$df[[i]]
mout3$df[[i]]
```
The two approaches should give almost equivalent results most of the time, but the new way is a little better.
